---
title: "Greenness_and_terrestrial_analyses"
author: "Matthew Murphy"
date: "10/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r lib_load}
# tools to generate phylogenetic tree
library(rotl)
library(ape)
library(phytools)
library(phylolm)
library(geiger)

# tool(s) for stat analysis
library(lawstat)
library(PMCMRplus)

# tool(s) to plot models
library(vioplot)
library(ggplot2)
library(latex2exp)

# piping and tidy
library(magrittr)
#library(gtools)
```

```{r }
load(file = 'dropped_species_ottid')
```
       
```{r }
# make lambda-max expression
l_max <- expression(lambda['max'])
```
## Make phylogenetic pipeline function

This is the pipeline that I used for creating the phylogenetic trees used in my analyses.
It requires the ape, phytools, and geiger packages.

```{r phyloPipeline}
phyloPipeline <- function(phy, dataframe, outgroup){
  origTree <- ape::read.tree((file=phy))
  treeTipBind <- phytools::bind.tip(origTree,
                                    tip.label = outgroup,
                                    edge.length = NULL,
                                    where = NULL,
                                    position = 0)
  treeRooted <- ape::root(phy = treeTipBind,
                          outgroup = outgroup)
  treeWithBranchLengths <- ape::compute.brlen(treeRooted)
  treeDichotomous <- ape::multi2di(treeWithBranchLengths)
  missing <- geiger::name.check(phy = treeDichotomous,
                                data = dataframe)
  treeFinal <- ape::drop.tip(phy = treeDichotomous,
                             tip = missing$tree_not_data)
  
  return(treeFinal)
}
```

## Import datasets

I manually reclassified and coded the (binary) habitat data.
This data set lacks the invertebrate/vertebrate binary.
```{r import_data}
dir = "~/Research/Data/meta-analysis/finalized_data/"
setwd(dir = dir)
aquatic <- read.csv(file = '~/Research/Data/meta-analysis/finalized_data/aquatic_final.csv', 
                    header = T, 
                    stringsAsFactors = T)
terrestrial <- read.csv(file = 'terrestrial_final.csv', 
                        header = T, 
                        stringsAsFactors = T)
otts_excluded <- read.csv(file = '~/Research/Data/meta-analysis/excluded_ottids_processed.csv', 
                          header = T,
                          stringsAsFactors = T)
otts_excluded$d_min <- as.numeric(otts_excluded$d_min)
otts_excluded$d_max <- as.numeric(otts_excluded$d_max)
otts_excluded$d_range <- as.numeric(otts_excluded$d_range)
rownames(aquatic) <- aquatic[["species_ottid"]]
rownames(terrestrial) <- terrestrial[["species_ottid"]]
rownames(otts_excluded) <- otts_excluded[["species_ottid"]]
```
## Adding excluded otts back in to data
```{r reAddOTTs}
d_avg <- (otts_excluded[,'d_min'] + otts_excluded[,'d_max'])/2
otts_excluded <- cbind(otts_excluded[,c(1:6)],d_avg,otts_excluded[,c(7:21)])

terrestrial_o <- rbind(terrestrial, otts_excluded[which(otts_excluded$habitat == 'forest'),])
aquatic_o <- rbind(aquatic, otts_excluded[-4,])
```

## Remove NAs from data

Phylogenetic analyses do not work correctly when there are NAs in the dummy variables.
I will need to remove them before starting phylogenetically controlled work

```{r removeNAs}
aquatic_RowsWithoutHabitatData <- which(is.na(aquatic$limnetic))
aquatic_naless <- aquatic[-aquatic_RowsWithoutHabitatData, ]

dummyVariables_Summed <- terrestrial[,c('forest','intermediate','open')] %>% apply(MARGIN = 1, FUN = sum)
            
dummyVariables_naless <- which(dummyVariables_Summed >= 0 )

terrestrial_naless <- terrestrial[dummyVariables_naless,]

rm(aquatic_RowsWithoutHabitatData, dummyVariables_Summed, dummyVariables_naless)
```

## Create combined dataset (aquatic and terrestrial)
I combined the two data sets into one, which has all dummy variables

```{r cbindterrestrial_aquaticfixed}
aquaterra <- rbind(aquatic_naless, terrestrial_naless)
```

### Get standard variables
```{r depvars}

# creating a vector of commonly used dependent variable names 
# allowing the user to iterate across them

dependent_variables <- c('o_max','o_min','o_range')
```


## Aquatic data

### Exploratory statistics
```{r aquaticExploratoryStats}

for(d in dependent_variables){
  print(d)
  aquatic[[d]] %>% hist(main = paste('Histogram of', d, 'include dropped otts', sep = ' '))
  aquatic[[d]] %>% shapiro.test() %>% print()
  
  aquatic_naless[[d]] %>% hist(main = paste('Histogram of', d, 'NA-less', sep = ' '))
  aquatic_naless[[d]] %>% shapiro.test() %>% print()
}
```

#### Pretty Depth Plots
```{r }
ylabels <- c('Longest', 'Shortest', 'Range of')
xlabels <- c('Minimum', 'Maximum', 'Average')
independent_variables <- c('d_avg','d_min','d_max')
for(i in independent_variables){
  iter <- 1
  for(d in dependent_variables){
    ylabel <- paste0(ylabels[[iter]], ' ', "$\\lambda_{max}$") #LaTeX y-axis label
    data <- aquatic_o[,c(i,d)] # Get columns to use for printing graph
    data <- data[complete.cases(data),] # Remove NAs from graph (won't print correctly otherwise)

    p <- ggplot(data = NULL, aes(x = data[[i]], y = data[[d]])) +
      geom_point(alpha = 0.75, color = 'black') +
      theme_classic() + # No gray background or graph grid
      {if(d != 'o_range')scale_y_continuous(limits = c(200, 500), breaks = seq(200,500,50))} +
      {if(d == 'o_range')scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, 25))} +
      {if(i == 'd_min')scale_x_continuous(limits = c(0, 500), breaks = seq(0, 500, 100))} +
      {if(i == 'd_max')scale_x_continuous(limits = c(0, 4500), breaks = seq(0, 4500, 500))} +
      {if(i == 'd_avg')scale_x_continuous(limits = c(0, 2500), breaks = seq(0, 2500, 500))} +
  # print axis labels
      {if(i == 'd_avg')labs(x = 'Average Depth  (m)')} +
      {if(i == 'd_min')labs(x = 'Minimum Depth  (m)')} +
      {if(i == 'd_max')labs(x = 'Maximum Depth  (m)')} +
      labs(y = TeX(ylabel)) +
  # calculate and draw regression line
      geom_smooth(method = glm, se = F)

  # visualization in HTML
      print(p)
      print(paste0('formula =', i, '~', d))
      print(summary(glm(formula = data[[i]] ~ data[[d]])))
  # save copy to disk
      file <- paste0(dir,'/','Depth_',d,'_by_',i,'include-dropped-otts.svg')
      ggsave(file = file, device = 'svg')
      iter <- iter + 1
  }
}
# cleanup
rm(i, d, iter, xlabels, ylabels, ylabel, file, data)
```

### Phylolm for visual sensitivities as a function of depth
```{r getPhylogenyForDepths}
phylogenies <- c('phyDepthMin_Final', 'phyDepthMax_Final', 'phyDepth_Final')

r <- 1
for(i in independent_variables){
  aquatic_RowsWithoutDepth <- which(is.na(aquatic_naless[,i]))
  depth_naless <- aquatic_naless[-aquatic_RowsWithoutDepth, ]
  
  finalDepthTree <- phyloPipeline(phy = 'test_phylo2_093021',
                                  dataframe = depth_naless,
                                  outgroup = 'Saccharomyces_cerevisiae_ottid5262624')
  assign(x = phylogenies[r], value = finalDepthTree)
  assign(x = paste0('aquatic_', i, '_naless'), value = depth_naless)
  
  for(d in dependent_variables){
  print(summary(phylolm(depth_naless[[d]] ~ depth_naless[[i]],
                        data = depth_naless,
                        phy = finalDepthTree,
                        boot = 100)))
  }
  r <- r+1
  rm(aquatic_RowsWithoutDepth, depth_naless, finalDepthTree)
}
```


## Terrestrial data
### Exploratory statistics
```{r}
for(d in dependent_variables){
  cat(d)
  terrestrial[[d]] %>% hist(main = paste('Histogram of', d, sep = ' '))
  terrestrial[[d]] %>% shapiro.test() %>% print()
  
  terrestrial_o[[d]] %>% hist(main = paste('Histogram of', d, 'no dropped OTTs', sep = ' '))
  terrestrial_o[[d]] %>% shapiro.test() %>% print()  

  cat('NO NAs')
  terrestrial_naless[[d]] %>% hist(main = paste('Histogram of', d, 'NA-less', sep = ' '))
  terrestrial_naless[[d]] %>% shapiro.test() %>% print()
  
  cat('\n\n\n')
  }
```

```{r terrestrialNMeanSD}
habitat <- unique(terrestrial_naless[,'habitat'])
levels(terrestrial_naless[,'habitat']) <- c('closed.intermediate','open.specialist','terrestrial.generalist')
terrestrial_o[which(terrestrial_o$habitat == 'forest'),'habitat'] <- 'closed.intermediate'
terrestrial_o[,'habitat'] <- as.character(terrestrial_o[,'habitat'])
terrestrial_o[,'habitat'] <- as.factor(terrestrial_o[,'habitat'])
levels(terrestrial_o[,'habitat']) <- c('closed.intermediate','open.specialist','terrestrial.generalist')

for(d in dependent_variables){
  print(d)

  sdev <- with(terrestrial_naless,
               aggregate(terrestrial_naless[,d],
                         by = list(category = habitat),
                         FUN = sd))
  m <- with(terrestrial_naless,
            aggregate(terrestrial_naless[,d],
                      by = list(category = habitat),
                      FUN = mean))
  summaryStats <- cbind(as.data.frame(habitat),m$x, sdev$x)
  colnames(summaryStats) <- c(paste0(d, ': habitat'),'mean','sdev')

  print(summaryStats)
  rm(sdev, m, summaryStats)
}
rm(d)
table(terrestrial_naless[,'habitat'])

print('INCLUDED DROPPED OTTS')
## without dropped otts
for(d in dependent_variables){
  print(d)

  sdev <- with(terrestrial_o,
               aggregate(terrestrial_o[,d],
                         by = list(category = habitat),
                         FUN = sd))
  m <- with(terrestrial_o,
            aggregate(terrestrial_o[,d],
                      by = list(category = habitat),
                      FUN = mean))
  summaryStats <- cbind(as.data.frame(habitat),m$x, sdev$x)
  colnames(summaryStats) <- c(paste0(d, ': habitat'),'mean','sdev')

  print(summaryStats)
  rm(sdev, m, summaryStats)
}
rm(d)
table(terrestrial_naless[,'habitat'])
print('INCLUDED DROPPED OTTS')
table(terrestrial_o[,'habitat'])
```
### Quick terrestrial plots
```{r quickTerrestrialPlots}
for(d in dependent_variables){
  plot(terrestrial_naless[[d]] ~ terrestrial_naless[['habitat']],
       xlab = 'habitat',
       ylab = d)
  plot(terrestrial_o[[d]] ~ terrestrial_o[['habitat']],
      xlab = 'habitat',
      ylab = d,
      main = 'INCLUDED DROPPED OTTS')
}
```
### Pretty terrestrial plots

For pretty terrestrial plots, see versions prior to 9-16-21

### Statistical analyses for terrestrial data
```{r terrstrialStats}
for(d in dependent_variables){
  kruskal <- (kruskal.test(terrestrial_naless[,d] ~ terrestrial_naless[,'habitat']))
  cat(paste('KRUSKAL-WALLIS TEST: \n EFFECT OF HABITAT ON:', d, sep = ' '))
  print(kruskal)
  cat('\n\n\n')
  
    kruskal <- (kruskal.test(terrestrial_o[,d] ~ terrestrial_o[,'habitat']))
  cat(paste('KRUSKAL-WALLIS TEST: \n EFFECT OF HABITAT ON:', d, sep = ' '))
  cat('\n')
  cat(paste0('INCLUDING DROPPED OTTS'))
  print(kruskal)
  cat('\n\n\n')
}
```



## Combined
```{r }
aquaterra_o <- rbind(aquatic_o, terrestrial_o)

write.csv(aquaterra_o, file = '~/Research/Data/meta-analysis/finalized_data/aquaterra_with_dropped_otts', row.names = F)
```
### Remove habitat generalists from analysis
```{r removeGeneralists}
# rows that describe generalists have NAs in the habitat column
aquaterra_generalist <- which(is.na(aquaterra$habitat))
aquaterra_specialist <- aquaterra[-aquaterra_generalist,]
rm(aquaterra_generalist)

aquaterra_generalist <- which(aquaterra_specialist$habitat == 'terrestrial.generalist')
aquaterra_specialist <- aquaterra_specialist[-aquaterra_generalist,]

aquaterra_specialist[,'habitat'] <- as.character(aquaterra_specialist[,'habitat'])
aquaterra_specialist[,'habitat'] <- as.factor(aquaterra_specialist[,'habitat'])

aquaterra_ogeneralist <- which(is.na(aquaterra_o$habitat))
aquaterra_ospecialist <- aquaterra_o[-aquaterra_ogeneralist,]
rm(aquaterra_ogeneralist)

aquaterra_ogeneralist <- which(aquaterra_ospecialist$habitat == 'terrestrial.generalist')
aquaterra_ospecialist <- aquaterra_ospecialist[-aquaterra_ogeneralist,]

aquaterra_ospecialist[,'habitat'] <- as.character(aquaterra_ospecialist[,'habitat'])
aquaterra_ospecialist[,'habitat'] <- as.factor(aquaterra_ospecialist[,'habitat'])
```
### Reimport prebuilt phylogenetic tree
```{r generate_tree}

aquaTerraTree <- phyloPipeline(phy = 'test_phylo2_093021',
                                     dataframe = aquaterra_specialist,
                                     outgroup = 'Saccharomyces_cerevisiae_ottid5262624')

# plot.phylo(x = aquaTerraTree, type = 'fan', cex = 0.1)
```

### Drop missing phylogenetic data
```{r dropmissing_tips_data}
missing <- geiger::name.check(phy = aquaTerraTree, data = aquaterra)

# exclude data not in tree
 exclude_rows <- which(row.names(aquaterra) %in% missing$data_not_tree)
 aquaterraFinal <- aquaterra_specialist # [-exclude_rows,]

 rm(missing, exclude_rows)
```

### Exploratory statistics
```{r}
for(d in dependent_variables){
  hist(aquaterraFinal[[d]],
       main = paste("Histogram of", d, sep = ' '),
       xlab = paste(d))
  print(paste('DEPENDENT VARIABLE:', toupper(d), sep = ' '))
  shatest <- (shapiro.test(aquaterraFinal[[d]]))
  print(shatest)
  
  print('INCLUDING DROPPED OTTS')
  hist(aquaterra_ospecialist[[d]],
       main = paste("Histogram of", d, ', including dropped otts', sep = ' '),
       xlab = paste(d))
  print(paste('DEPENDENT VARIABLE:', toupper(d), sep = ' '))
  shatest <- (shapiro.test(aquaterra_o[[d]]))
  print(shatest)
}
```

```{r aquaterraNMeanSD}

for(d in dependent_variables){
  sdev <- with(aquaterra_specialist,
               aggregate(aquaterra_specialist[,d],
                         by = list(category = habitat),
                         FUN = sd))
  colnames(sdev) <- c(paste('DEPENDENT VARIABLE:',toupper(d),sep=' '),'SDEV')
  m <- with(aquaterra_specialist,
            aggregate(aquaterra_specialist[,d],
                      by = list(category = habitat),
                      FUN = mean))
  final <- cbind(sdev,m[,2])
  colnames(final)[3] <- 'MEAN'
  print(sdev)
  print(m)
  print(final)
#  rm(sdev, m, final)
  
  print('INCLUDING DROPPED OTTS')
    sdev <- with(aquaterra_ospecialist,
               aggregate(aquaterra_ospecialist[,d],
                         by = list(category = habitat),
                         FUN = sd))
  colnames(sdev) <- c(paste('DEPENDENT VARIABLE:',toupper(d),sep=' '),'SDEV')
  m <- with(aquaterra_ospecialist,
            aggregate(aquaterra_ospecialist[,d],
                      by = list(category = habitat),
                      FUN = mean))
  final <- cbind(sdev,m[,2])
  colnames(final)[3] <- 'MEAN'
  print(sdev)
  print(m)
  print(final)
#  rm(sdev, m, final)
}

cat('SAMPLE SIZES')
table(aquaterra_specialist[,'habitat'])
cat('including dropped otts')
table(aquaterra_ospecialist[,'habitat'])
```

### test for equality of variances
```{r }
for(d in dependent_variables){
  print(levene.test(aquaterra_specialist[,d], group = aquaterra_specialist[,'habitat']))
  
  print('including droppsed otts')
  
  print(levene.test(aquaterra_ospecialist[,d], group = aquaterra_ospecialist[,'habitat']))
}

```

### Get 5 number summaries, means and SDs
#### need to incorporate?
```{r mean_sd}
aggregate(formula = o_max ~ habitat, data = aquaterra_ospecialist, FUN = 'mean')
aggregate(formula = o_max ~ habitat, data = aquaterra_ospecialist, FUN = 'sd')

aggregate(formula = o_min ~ habitat, data = aquaterra_ospecialist, FUN = 'mean')
aggregate(formula = o_min ~ habitat, data = aquaterra_ospecialist, FUN = 'sd')

aggregate(formula = o_range ~ habitat, data = aquaterra_ospecialist, FUN = 'mean')
aggregate(formula = o_range ~ habitat, data = aquaterra_ospecialist, FUN = 'sd')
# print('longest')
# print(tapply(green.open$o_max, green.open$habitat, summary))
```

### Roughly visualize data
```{r rough_viz}
for(d in dependent_variables){
  plot(aquaterra_specialist[,d] ~ aquaterra_specialist[,'habitat'],
       ylab = paste(d),
       xlab = 'Habitat')
  
  plot(aquaterra_ospecialist[,d] ~ aquaterra_ospecialist[,'habitat'], main = 'including dropped otts',
       ylab = paste(d),
       xlab = 'Habitat')
}
```

### Pairwise tests
```{r kruskalDwass}
for(d in dependent_variables){
  print(d)
  response <- aquaterra_specialist[,d]
  group <- aquaterra_specialist[, "habitat"]
  kruskal_result <- kruskal.test(response ~ group)
  print(kruskal_result)
  if(kruskal_result$p.value<=0.05){
    dwass_result <- PMCMRplus::dscfAllPairsTest(response ~ group)
    print(dwass_result)
#    print(summary(dwass_result))
    print(dwass_result$statistic)
  }
  rm(response,group,kruskal_result,dwass_result)
  
  print('including dropped otts')
  response <- aquaterra_ospecialist[,d]
  group <- aquaterra_ospecialist[, "habitat"]
  kruskal_result <- kruskal.test(response ~ group)
  print(kruskal_result)
  if(kruskal_result$p.value<=0.05){
    dwass_result <- PMCMRplus::dscfAllPairsTest(response ~ group)
    print(dwass_result)
 #   print(summary(dwass_result))
    print(dwass_result$statistic)
  }
  # rm(response,group,kruskal_result,dwass_result)
}
```

### Correct by phylogeny

```{r phylolm_max_terrestrialaquatic}
print(summary(phylolm(o_max ~ coastal + lake.river + open,
                      data = aquaterraFinal,
                      phy = aquaTerraTree,
                      boot =100)))

# get open
print(summary(phylolm(o_max ~ coastal + lake.river + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))

# get coastal
print(summary(phylolm(o_max ~ open + fresh + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))
```

```{r phylolm_min_terrestrialaquatic}
print(summary(phylolm(o_min ~ coast + fresh + open,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))

# get open
print(summary(phylolm(o_min ~ coast + fresh + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))

# get coastal
print(summary(phylolm(o_min ~ open + fresh + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))
```

```{r phylolm_range_terrestrialaquatic}

print(summary(phylolm(o_range ~ coast + fresh + open,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))

# get open
print(summary(phylolm(o_range ~ coast + fresh + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))

# get coastal
# get open
print(summary(phylolm(o_range ~ open + fresh + closed_int,
                      data = green.open,
                      phy = green.open.brx,
                      boot =100)))
```